The Second Data Release of the INT Photometric H$\alpha$ Survey of the Northern Galactic Plane (IPHAS DR2)

### 2014-04-14 paper/iphas-dr2.tex
Clarified that seeing really means PSF FWHM

(1) \emph{Depth.} 
We discarded any exposures for which the $5\sigma$ limiting magnitude 
was worse than 20th magnitude in the $r$-band
or worse than 19th in $i$ or \ha. 
Such data were typically obtained during poor weather or full moon.
Most observations were significantly better than these limits.
Fig.~\ref{fig:depth} presents the distribution of limiting magnitudes
for all quality-approved fields,
showing a mean depth of 
$21.2\pm0.5$ ($r$), $20.0\pm0.3$ ($i$) and $20.3\pm0.3$ (\ha).
The depth achieved depended 
most strongly on the presence of the moon,
which was above the horizon during 62\% 
of the observations.  The great range in sky brightness this
produced is the reason for the wide and bi-modal shape
of the $r$-band limiting magnitude distribution 
(top panel in Fig.~\ref{fig:depth}).
In contrast, the depths attained in $i$ and \ha\ 
are less sensitive to moonlight, leading to
narrower magnitude limit distributions
(middle and bottom panels in Fig.~\ref{fig:depth}).

The minimum depth criterion
led us to exclude 9\% of the pipelined data,
which had typically been obtained under cloudy conditions.
We note that some of the excluded data may nevertheless be useful
for e.g. time-domain studies of bright stars.
The detection tables of the discarded data are made
available through our website for this reason (www.iphas.org),
but are ignored in what follows.

(2) \emph{Ellipticity.} 
The ellipticity of a point source,
defined as $e = 1 - b / a$ 
with $b$ the semi-minor and $a$ the semi-major axis,
is a morphological measure of the elongation of the point spread function.
It is expected to be zero (circular) across the field 
in a perfect imaging system,
but it is slightly non-zero in any real telescope data 
due to optical distortions and tracking errors.  Indeed, it is worth noting
that IPHAS data have been collected from unguided exposures that rely
entirely on the INT's tracking capability.
The mean ellipticity measured 
%across a field 
from exposures forming this release
%in the IPHAS data set 
is $0.09\pm0.04$.
There have been sporadic episodes with higher ellipticities
due to mechanical glitches in the telescope tracking system.
To exclude these, we rejected exposures in which the mean ellipticity
across the detector exceeded $e > 0.3$. We have also found that, as
this threshold is breached, the photometric measurements delivered by 
the pipeline become degraded.
Only 0.4\% of the exposures were discarded on this basis.

(3) \emph{Seeing.} 
The original survey goal was to obtain data 
at seeing better than 1.7~arcsec.
This target is currently attained across 86\% of the footprint,
in particular at lower Galactic longitudes,
e.g. 92\% of the fields at $l<120^\circ$ are better than 1.7~arcsec.
Fig.~\ref{fig:seeing} presents the distribution
of the mean seeing for all the quality-approved fields.
We find a median value of 1.1~arcsec in $r$, \ha\
and 1.0~arcsec in $i$.
In the $r$-band, 90\% of the data is better than 1.5~arcsec,
and 10\% is better than 0.8~arcsec.
To increase the sky area covered by this data release,
we have decided to include the small fraction of data
obtained under seeing of up to 2.5 arcsec.  This limited
exclusion on grounds of poor seeing to only 1\%\ of the pipelined 
exposures. 
In \S\ref{sec:catalogue} we will explain
that the photometry listed in the catalogue
is predominantly based on the images with the
best-available seeing, and that the seeing estimate
is included as a column in the catalogue.

\begin{figure}
    \begin{minipage}[b]{\linewidth}
        \includegraphics[width=\textwidth]{./plots/seeing_r.pdf} 
    \end{minipage}
    \begin{minipage}[b]{\linewidth}
        \includegraphics[width=\textwidth]{./plots/seeing_i.pdf} 
    \end{minipage}
    \begin{minipage}[b]{\linewidth}
        \includegraphics[width=\textwidth]{./plots/seeing_ha.pdf} 
    \end{minipage}
    \caption{Seeing distribution across all
             quality-approved survey fields
             for $r$ (top), $i$ (middle) and \ha\ (bottom).}
    \label{fig:seeing}
\end{figure}

(4) \emph{Photometric repeatability.} 
The IPHAS field-pair observing strategy normally
ensures that every pointing is immediately followed 
by an offset pointing at a displacement of $+$5~arcmin in Dec 
and $+$5~arcmin in RA.
This allows pairs of images to be checked 
for the presence of clouds or electronic noise.
To exploit this information,
the overlap regions of all field pairs were systematically cross-matched
to verify the consistency of the photometry
for stars observed in both pointings.
We rejected field pairs in which more than 2\% of the stars 
showed an inconsistent measurement at the level of 0.2 mag,
or more than 26\% were inconsistent at the level of 0.1 mag.
These limits were set empirically after inspecting
the images and photometry by eye.
We rejected 11\% of the data as part of this step
(many of which had already been rejected as part of the limiting magnitude criterion).

(5) \emph{Source density mapping.}
Spatial maps showing the number density of the detected sources
down to 20th magnitude were created to verify the health
of the data and to check for unexpected artefacts.
%(e.g. bright satellite trails)
In particular, we created density maps
which showed the number of \emph{unique} sources
obtained by cross-matching the detection tables of
all three bands with a maximum matching distance of 1~arcsec.
This was particularly effective for revealing
fields with an inaccurate astrometric solution in one of the bands.
In the majority of the cases we were able to correct the astrometry
of such fields by hand.

(6) \emph{Visual examination.}
All images and their associated photometric colour/magnitude diagrams
were inspected by a team of 20 survey members, 
such that each image in the data release 
was looked at by at least three different pairs of eyes.
Images deemed unsuitable were flagged, investigated and
excluded from the release if necessary. 
6\% of the observations were placed on a \emph{black-list}
as a result of this -- most commonly because of the obvious 
presence of clouds or extreme levels of scattered moonlight.

(7) \emph{Contemporaneous field data.} 
Finally, only exposures which are part of a sequence 
of three consecutive images of the same field
(\ha, $r$, $i$)
were considered for inclusion in the release. 
This ensures that the three bands for a given field
are observed contemporaneously --  
typically within 5 minutes of each other.
An exception was made for 9 fields where the three exposures 
could not be obtained within the same night
but for which the time gap between exposures did not exceed 48 hours.
We note that the source catalogue details the exact epoch
at the start of each exposure
(columns \emph{rMJD}, \emph{iMJD}, \emph{haMJD}).

All of the above criteria were satisfied by at least one observing attempt
for 14\,115 out of the 15\,270 fields (92\%).

In some cases more than one successful attempt to observe
a field was available due to stricter
quality criteria being applied in the initial years of the survey.
In such cases, only the attempt 
with the best seeing and depth has been selected
for inclusion in the catalogue, in order  
to deliver the most reliable measurement at a single epoch.
Those interested in any of the rejected data 
can nevertheless data-mine the full set of detection tables 
via our website.

@@

   pStar $>$ 0.9 &
   145 million (66\%) &
   % pStar > 0.89 is identical to mergedClass == -1
   The object must appear as a perfect point source,
   as inferred from comparing its Point Spread Function (PSF)
   with the average PSF measured in the same CCD. \\

>>>>

(1) \emph{Depth.} 
We discarded any exposures for which the $5\sigma$ limiting magnitude 
was worse than 20th magnitude in the $r$-band
or worse than 19th in $i$ or \ha. 
Such data were typically obtained during poor weather or full moon.
Most observations were significantly better than these limits.
Fig.~\ref{fig:depth} presents the distribution of limiting magnitudes
for all quality-approved fields,
showing mean depths of 
$21.2\pm0.5$ ($r$), $20.0\pm0.3$ ($i$) and $20.3\pm0.3$ (\ha).
The depth achieved depended 
most strongly on the presence of the moon,
which was above the horizon during 62\% 
of the observations.  The great range in sky brightness this
produced is the reason for the wide and bi-modal shape
of the $r$-band limiting magnitude distribution 
(top panel in Fig.~\ref{fig:depth}).
In contrast, the depths attained in $i$ and \ha\ 
are less sensitive to moonlight, leading to
narrower magnitude limit distributions
(middle and bottom panels in Fig.~\ref{fig:depth}).

(2) \emph{Ellipticity.} 
The ellipticity of a point source,
defined as $e = 1 - b / a$ 
with $b$ the semi-minor and $a$ the semi-major axis,
is a morphological measure of the elongation of the point spread function (PSF).
It is expected to be zero (circular) across the field 
in a perfect imaging system,
but it is slightly non-zero in any real telescope data 
due to optical distortions and tracking errors.
Indeed, it is worth noting
that IPHAS data have been collected
from unguided exposures that rely entirely
on the INT's tracking capability.
The mean ellipticity measured
in the data is $0.09\pm0.04$.
There have been sporadic episodes with higher ellipticities
due to mechanical glitches in the telescope tracking system.
To exclude these, we rejected exposures in which the mean ellipticity
across the detector exceeded $e > 0.3$.
We have also found that, as this threshold is breached,
the photometric measurements delivered by the pipeline become degraded.

(3) \emph{Seeing.} 
The original survey goal was to obtain data 
at a resolution better than 1.7~arcsec,
as evaluated by measuring the average PSF Full Width at Half Maximum (FWHM)
across an image,
which is a proxy for the combined effect of atmospheric and dome seeing.
This target is currently attained across 86\% of the footprint,
in particular at lower Galactic longitudes,
e.g. 92\% of the fields at $l<120^\circ$ are better than 1.7~arcsec.
To slightly increase the sky area offered by this data release,
we have decided to include data obtained under seeing of up to 2.5 arcsec.
Fig.~\ref{fig:seeing} presents the distribution
of the PSF FWHM for the approved fields in the release.
In the $r$-band, 90\% of the data is better than 1.5~arcsec,
50\% is better than 1.1~arcsec,
and 10\% is better than 0.8~arcsec.
In \S\ref{sec:catalogue} we will explain
that the photometry compiled in the source catalogue
is normally derived from the field with the
best-available seeing for a given object,
and that the PSF FWHM measurement
is included as a column in the catalogue.

\begin{figure}
    \begin{minipage}[b]{\linewidth}
        \includegraphics[width=\textwidth]{figures/seeing/seeing_r.pdf} 
    \end{minipage}
    \begin{minipage}[b]{\linewidth}
        \includegraphics[width=\textwidth]{figures/seeing/seeing_i.pdf} 
    \end{minipage}
    \begin{minipage}[b]{\linewidth}
        \includegraphics[width=\textwidth]{figures/seeing/seeing_ha.pdf} 
    \end{minipage}
    \caption{Distribution of the PSF FWHM
             for all the quality-approved fields
             included in the release,
             measured in $r$ (top), $i$ (middle) and \ha\ (bottom).
             The PSF FWHM measures the effective image resolution
             that arises from the combination of atmospheric and dome seeing.}
    \label{fig:seeing}
\end{figure}

(4) \emph{Photometric repeatability.} 
The IPHAS field-pair observing strategy normally
ensures that every pointing is immediately followed 
by an offset pointing at a displacement
of $+$5~arcmin in Dec and $+$5~arcmin in RA.
This allows pairs of images to be checked 
for the presence of clouds or electronic noise.
To exploit this information,
the overlap regions of all field pairs were systematically cross-matched
to verify the consistency of the photometry
for stars observed in both pointings.
We rejected field pairs in which more than 2\% of the stars 
showed an inconsistent measurement at the level of 0.2 mag,
or more than 26\% were inconsistent at the level of 0.1 mag.
These limits were set empirically after inspecting
the images and photometry by eye.

(5) \emph{Source density mapping.}
Spatial maps showing the number density of the detected sources
down to 20th magnitude were created to verify the health
of the data and to check for unexpected artefacts.
In particular, we created density maps
which showed the number of \emph{unique} sources
obtained by cross-matching the detection tables of
all three bands with a maximum matching distance of 1~arcsec.
This was particularly effective for revealing
fields with an inaccurate astrometric solution in one of the bands,
which were subsequently corrected.

(6) \emph{Visual examination.}
Images, colour mosaics,
and the associated photometric colour/magnitude diagrams
were inspected by a team of 20 survey members, 
such that each image in the data release 
was looked at by at least three different pairs of eyes.
Images affected by clouds or extreme levels of scattered moonlight
were flagged, investigated,
and excluded from the release 
by placing them on a \emph{black-list}.

(7) \emph{Contemporaneous field data.} 
Finally, only exposures which are part of a sequence 
of three consecutive images of the same field
(\ha, $r$, $i$)
were considered for inclusion in the release. 
This ensures that the three bands for a given field
are observed contemporaneously --  
normally within 5 minutes of each other.
An exception was made for 9 fields where the three exposures 
could not be obtained within the same night
but for which the time gap between exposures did not exceed 48 hours.
We note that the source catalogue details the exact epoch
at the start of each exposure
(columns \emph{rMJD}, \emph{iMJD}, \emph{haMJD}).

All of the above criteria were satisfied by at least one observing attempt
for 14\,115 out of the 15\,270 fields (92\%).
In some cases more than one successful attempt to observe
a field was available due to stricter
quality criteria being applied in the initial years of the survey.
In such cases, only the attempt 
with the best seeing and depth has been selected
for inclusion in the catalogue, in order  
to deliver the most reliable measurement at a single epoch.

We note that some of the excluded data may nevertheless be useful
for e.g. time-domain studies of bright stars.
The images and detection tables of the discarded data are made
available through our website for this reason (www.iphas.org),
but will be ignored in the remainder of this work.

@@

   pStar $>$ 0.9 &
   145 million (66\%) &
   % pStar > 0.89 is identical to mergedClass == -1
   The object must appear as a perfect point source,
   as inferred from comparing its PSF
   with the average PSF measured in the same CCD. \\
